---
title: "DESeq2 analysis"
author: "Bartosz Bartmanski"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tximport)
library(DESeq2)
library(here)


get_contrast <- function(strain, comp) {
    return(c(
        "group", 
        paste0(strain, "_", str_sub(comp, 1, 1)), 
        paste0(strain, "_", str_sub(comp, 4, 4))
    ))
}
```

## Load data

First, we create a dataframe of files that need to be imported and 
corresponding columns for strain id, condition, replicate, which we can get 
from the file names.

```{r}
quant_files <- list.files(here("Output/Quants/"), recursive = T, pattern = "quant.sf", full.names=T)
samples <- quant_files %>% dirname() %>% basename()

df_samples <- data.frame(
    row.names=samples, 
    sample=samples, 
    files=quant_files
) %>%
separate_wider_delim(
    cols="sample", 
    delim="_", 
    names=c("strain", "condition", "replicate"), 
    cols_remove=F
)

selected_strains <- df_samples %>% pull(strain) %>% unique()
```

Next, we use the tximport package to load the output from salmon for all samples
in a format DESeq2 expects:

```{r pressure, echo=FALSE}
# Read in the annotation file
tx2gene <- read_tsv(here("Data/annotation.tsv"), show_col_types=F) %>% 
select(transcript_stable_id, gene_stable_id)

# Create metadata table
meta <- df_samples %>%
mutate(group=factor(paste(strain, condition, sep="_"))) %>%
select(sample, condition, group) %>% 
column_to_rownames(var="sample")

# Import all the salmon outputs
txi <- tximport(
    df_samples[["files"]], 
    type="salmon", 
    tx2gene=tx2gene, 
    countsFromAbundance="lengthScaledTPM"
)
```
## Running DESeq2

```{r}
# Run DESeq2 
dds <- DESeqDataSetFromTximport(txi, colData = meta, design = ~ group)
dds <- DESeq(dds)

# Get raw counts and save to csv
df_counts <- dds %>%
counts() %>% 
as.data.frame() %>% 
rownames_to_column(var="gene_id")

# Get normalised counts and save to csv
df_norm <- dds %>%
counts(normalized=TRUE) %>% 
as.data.frame() %>% 
rownames_to_column(var="gene_id")

# and save to rds
dir.create(here("Output/DESeq2/"), showWarnings = F)
saveRDS(dds, here("Output/DESeq2/dds.rds"))
```

Finally, we get the log fold changes and p-values for each comparison:

```{r}
df_lfc <- expand_grid(
    strain=selected_strains, 
    comparison=c("MvsP", "CvsP", "CvsM")
) %>% 
# filter(!(strain == "090.3" & str_detect(comparison, "M"))) %>%
rowwise() %>%
mutate(both = list(c_across(everything()))) %>%
pull(both) %>%
map(
    ~{
        results(dds, contrast=get_contrast(.x[[1]], .x[[2]])) %>%
        as.data.frame() %>%
        rownames_to_column(var="gene_id") %>%
        mutate(strain=.x[[1]], comparison=.x[[2]])
    }
) %>%
bind_rows() %>%
dplyr::rename(log2fc=log2FoldChange) %>%
mutate(regulation = case_when(padj < 0.05 & log2fc > 1 ~ "Up",
                              padj < 0.05 & log2fc < -1 ~ "Down",
                              TRUE ~ "None"))

df_lfc %>% write_csv(here("Output/DESeq2/dds_results.csv"))
```

## Plotting results

```{r}
volcano_plot <- function(data, pval=0.05, lfcval=1, col="padj") {
    ggplot(data=data, 
           aes(x=log2fc, y=-log10(!!sym(col)), col=regulation, gene_id=gene_id)) +
        geom_point(size=0.5) + 
        scale_color_manual(values=c("Up"="red", "Down"="blue", "None"="black")) +
        geom_hline(yintercept=-log10(pval), linetype='dashed', color="red") + 
        geom_vline(xintercept=-lfcval, linetype='dashed', color="red") +
        geom_vline(xintercept=lfcval, linetype='dashed', color="red") %>%
    return()
}
```

Volcano plot for strain 083.2 comparing M treatment and parental strain:

```{r}
df_lfc %>%
filter(strain == "083.2" & comparison == "MvsP") %>%
volcano_plot()  # %>% plotly::ggplotly()
```
All volcano plots:

```{r}
df_lfc %>%
volcano_plot() +
facet_grid(col=vars(strain), row=vars(comparison)) 
```
Number of significantly up- or down-regulated genes:

```{r}
df_lfc %>%
filter(regulation != "None") %>% 
dplyr::count(strain, comparison, regulation) %>%
ggplot(aes(x=strain, y=n, fill = regulation)) +
    geom_bar(stat="identity", position="dodge", width=0.5) +
    scale_fill_manual(values=c("blue", "red")) +
    facet_grid(rows=vars(comparison))
```

## Genes in common between strains

Number of genes, that are significantly up- or down-regulated that are common 
between strains:

```{r}
df_num_strains <- df_lfc %>% 
filter(padj < 0.05 & abs(log2fc) > 1) %>%
group_by(comparison, gene_id) %>%
summarise(num_strains=n_distinct(strain)) %>%
ungroup() %>%
group_by(num_strains, comparison) %>%
summarise(num_genes=n_distinct(gene_id)) %>% 
ungroup()

df_num_strains %>%
ggplot(aes(x=num_strains, y=num_genes, fill=comparison)) +
    geom_bar(stat="identity", position="dodge", width=0.5)
```


## PCA

```{r}
# Get variance stabilising transformation of counts
df_vst <- dds %>% 
vst() %>% 
assay() %>% 
as.data.frame() %>%
rownames_to_column(var="gene_id")

pca <- df_vst %>% 
column_to_rownames(var="gene_id") %>% 
t() %>%
prcomp(retx=T, rank.=2)

df_pca <- pca[["x"]] %>% 
as.data.frame() %>%
rownames_to_column(var="group") %>%
tidyr::separate(group, into=c("strain", "condition", "replicate"), sep="_")

summary(pca)

var_expl <- summary(pca)$importance[2, c(1, 2)]
```

```{r}
df_pca %>%
ggplot(mapping=aes(x=PC1, y=PC2)) + 
    geom_point(size=2) + 
    labs(
        x=paste0("PC1 (", var_expl["PC1"]*100, "%)"),
        y=paste0("PC2 (", var_expl["PC2"]*100, "%)")
    )
```

**Task**: try adding strain, condition or MIC information into the plot to see 
any clustering.

The MIC table is loaded below:

```{r}
df_mic <- read_csv(here("Data/mic.csv")) %>%
mutate(
    cza_mic_level=ifelse(cza_mic > 8, "high", "low"), 
    mem_mic_level=ifelse(mem_mic > 8, "high", "low")
)
```

